R workflow
========================================================
preliminary exploration and analysis of Borneo survival data
--------------------------------------------------------
##### Sean Tuck - PhD student, Oxford
### Description
Workflow notes that distill the main points from accompanying code (Mortality_GLMM_Code.R) written to explore and begin modelling Borneo mortality analysis (MaluaDataI.txt).  
Binomial GLMMs will be fitted to data, to estimate mortality of species and assess the relative importance
of measured local environmental conditions in determining mortality.
Read metadata (Malua_MetadataDataI.xlsx) file and [Hector *et al.* 2011](http://rstb.royalsocietypublishing.org/content/366/1582/3303.full) for more information on data.  
##### Last updated 01/03/2013
***  
### Queries on bugs/fixes: 
1.  18 unique genus species combinations found (not incl. `Dont plant`). Plots contain max. of 16 species. Unique species found when combining genus and species info that are not included in list on MaluaPlantingDesign.doc, are:  
> 1. Dryobalanops tomentella
> 2. Hopea faguetiana

    If this should be corrected, which component of species info is incorrect - species or genus?
    The following code would preserve erroneous lines's species data and make the genera data conform to 
    relevant genera in MaluaPlantingDesign.doc, and then create a new file MaluaDataI_EDIT.txt 
    (rename MaluaDataI.txt if overwriting old file is needed):  
    
    ```
    MortDat$genusspecies<- as.factor(paste(MortDat$genus,MortDat$species,sep=" "))
    MortDat$genus[MortDat$genusspecies == 'Dryobalanops tomentella']<- factor('Parashorea')  
    MortDat$genus[MortDat$genusspecies == 'Hopea faguetiana']<- factor('Shorea')  
    MortDat<- MortDat[,1:32]  
    write.table(MortDat, file='MaluaDataI_EDIT.txt', sep='\t', row.names=FALSE)  
    ```

2.  The metadata file says "NA for all real missing values and unplanted points (1-9)".  
But `unique(MortDat$survival[MortDat$unpntdpoint != 0])` shows that there are 0s, not just NAs, at some unplanted points.

3.  Malua_MetadataDataI.xls shows 4 `canopycover` classes. In dataset, NA, 1-4, and 5 and 7 are found. Are 5 and 7 erroneous?

4.  NB: MaluaPlantingDesign.doc: `gd`:`fd` matrix shows one class in 4 `gd` with 2 `fd` (tall & medium III). The species listed underneath are all tall and the data show no 4 `gd` with 2 `fd` (see `table(MortDat$gd,MortDat$fd))`. Make change in PlantingDesign document?

***
### Data Exploration
Read in the data (currently my edited file - see code in queries section above), referred to in the text as `MortDat` and load `lme4` and `arm` packages.
```{r include=FALSE}
library(lme4)
library(arm)
MortDat<- read.table('MaluaDataI_EDIT.txt', header=TRUE, sep='\t')
```
Quick look at survival by species.
```{r fig.width=20}
spsurvival<- table(MortDat$species, MortDat$survival)
spsurvival
barplot(matrix(c(spsurvival[,1],spsurvival[,2]),ncol=length(spsurvival)), beside=TRUE, legend.text=c("Died","Survived"), ylab="Frequency", main="Numbers by species of survived and dead specimens", axis.lty="solid", cex.names=0.4, col=c('black','grey'))
```
Seems like there is a fair bit of variation between species, both in number of total counts, and dead/survived.

Do any species have a higher frequency of survived than dead? Let's express survival as a proportion too.
```{r}
which(spsurvival[,2] > spsurvival[,1])
spsurvival<- cbind(spsurvival,Proportion=spsurvival[,2]/(spsurvival[,1]+spsurvival[,2]))
summary(spsurvival[,3])
```
*Hopea sangal* is the only plant with more observed survivals than deaths.  
Observed species-averaged survival range from 15-60%, with mean 35%.

How many species are under different planting conditions?
```{r fig.width=20}
table(MortDat$species,MortDat$sd)
par(mfrow=c(1,3)); for(i in 1:length(unique(MortDat$sd))) hist(table(MortDat$species,MortDat$sd)[,i],xlab=NULL,main=sort(unique(MortDat$sd))[i]); par(mfrow=c(1,1)) 
```
Once `Dont plant` is accounted for, most species in 1`sd` and 16`sd` conditions have similar number of records.

Observed survival under different planting conditions (`sd`).
```{r}
spsdsurvival<- table(MortDat$species,MortDat$sd,MortDat$survival)
spsdsurvival
spsdproportion<- spsdsurvival[,,2]/(spsdsurvival[,,1]+spsdsurvival[,,2])
boxplot(spsdproportion)
```
Looks to be no great change between `sd` classes, maybe other than slightly less variation at high `sd`.
```{r}
plot(as.factor(sort(unique(MortDat$sd))),NULL,ylim=c(0.1,0.7),type='n')
cols<- rainbow(nrow(spsdproportion))
for(i in 1:nrow(spsdproportion)) lines(as.factor(sort(unique(MortDat$sd))),spsdproportion[i,], col=cols[i])
```
There seems to be no real trend in by-species proportion survived with `sd` classes. Some increase with `sd`, some decrease.
If anything, many are lowest at intermediate `sd`. But again, markedly different among species.

Survival vs. damage types.
```{r}
table(MortDat$insectdamage, MortDat$survival)
table(MortDat$mammaldamage, MortDat$survival)
table(MortDat$treefalldamage, MortDat$survival)
```
Not much increase in deaths as any type of damage increases in intensity.

Records for by-species by-plot survival, to look at the spread across plots where records are found for all species.
```{r}
tapply(MortDat$pl,MortDat$species,summary)
```
Distribution similar for all species.

Boxplot of `timelag` (`surveydate` - `plantingdate`) conditional on survival.
```{r fig.width=20}
boxplot(MortDat$timelag ~ MortDat$survival, ylab="timelag",xlab="survival", horizontal=TRUE)
```
Spread of `timelag` values similar for both survived and dead observations.

### Notes:
Based on data exploration, I would expect diversity (`sd`) to have little effect on estimated species mortality rates. Some species seem to be more likely to survive in mixture, and others when grown independently, but few responses seem dramatic. Observed proportion of survival differs greatly between species, so I expect estimated mortality rates to be different between species.

Looks to be no significant impact of `timelag` (time between planting and survey) on observing survival.

The number of species found under 1`sd` and 16`sd` conditions is relatively evenly spread. 4`sd` shows some greater variation, mainly because 4 species (*D. conformis*, *D. lanceolata*, *H. sangal*, and *H. spp.*) occur much more regularly than the rest. At 4`sd`, the least occurring species is *S. leprosula*, at 638 records. So, at 4`sd` at least, there is some variation in how well each species is represented.

The distributions of planting times between species are very similar - data were collected at similar times, so there is minimal species-associated bias or skew in collection dates (and by extension in any corresponding temporal-specific conditions).

***
### Modelling
`REML=FALSE` used so models can be compared by AIC when appropriate.
```{r}
ranefmod<- lmer(survival ~ 1 + (1|foresttype) + (1|pl), data=MortDat, family=binomial, REML=FALSE)
summary(ranefmod)
```
Null model - just random effects. Plot level random effect to deal with variation due to unexplained effects at plot level. Random intercept included to account for differences between forest types; if observations of habitat characteristics, such as canopy cover are used in the model as fixed effects, then it would be best to omit this. Likelihood Ratio Test shows inclusion of forest types random effect is a significant improvement.

How are effects of forest characteristics best specified? Do we want to estimate parameters for different forest types? Outputs of representing forest types as a random effect, and representing a characteristic of forest type (canopy cover) as a fixed effect are compared.
```{r}
spcnpymod<- lmer(survival ~ canopycover + (1|pl), data=MortDat, family=binomial, REML=FALSE)
spcnpymod2<- lmer(survival ~ canopycover + (1|foresttype) + (1|pl), data=MortDat, family=binomial, REML=FALSE)
anova(ranefmod,spcnpymod2)
AIC(ranefmod)-AIC(spcnpymod)
```
Canopy cover as a fixed effect is significant (although parameter estimate is close to 0), but becomes insignificant with forest type random intercept included. Forest type is capturing that variation attributable to forest types, and AIC comparison shows lower deviance when this variation is represented as a random effect. Specifying this as a random effect makes more sense anyway; estimating parameters for different densities of canopy cover will consider only one of the many ways in which different forest types can vary. The random effects structure will be maintained and canopy cover will be omitted.

Does including species as a fixed effect, estimating a parameter for each, provide additional explanatory power?
```{r}
spmortmod<- lmer(survival ~ species + (1|foresttype) + (1|pl), data=MortDat, family=binomial, REML=FALSE)
anova(ranefmod,spmortmod)
```
Again, Likelihood Ratio Test shows that estimating a mortality rate for each species provides significantly more explanatory power than estimating a species-average response, as expected from data exploration. Species categories will be kept in the model.

Does species-level diversity have a significant effect on mortality? Species diversity classes included as a fixed effect.
```{r}
spsdmortmod<- lmer(survival ~ species + sd + (1|foresttype) + (1|pl), data=MortDat, family=binomial, REML=FALSE)
anova(spmortmod,spsdmortmod)
```
This does not improve the models explanatory power and shows diversity to have no effect on the mortality of a species.

What about other levels of diversity? Does increased diversity at the genus level have a significant effect on mortality? What about varying levels of functional diversity (canopy mixture) nested within generic diversity? Try fitting some kind of interaction between functional diversity and generic diversity? How best to specify this?
```{r}
spgdfdmod<- lmer(survival ~ species + gd*fd + (1|foresttype) + (1|pl), data=MortDat, family=binomial, REML=FALSE)
spgdmortmod<- lmer(survival ~ species + gd + (1|foresttype) + (1|pl), data=MortDat, family=binomial, REML=FALSE)
spfdmortmod<- lmer(survival ~ species + fd + (1|foresttype) + (1|pl), data=MortDat, family=binomial, REML=FALSE)
```
No sign of a significant effect of increasing generic diversity, functional diversity in terms of canopy mixture, or including the interaction between increasing functional diversity as generic diversity increases.

#### NB:
Models that include timelag and types of damage are not converging on meaningful values. Changing the approximation technique from Laplacian to Gauss-Hermite Quadrature (greater accuracy) does not solve false convergence. Leads to wild model estimates. Models that include any size covariates (continuous variables) also fail to converge...
***
### Minimal adequate model
```{r}
spmortmod
binnedplot(fitted(spmortmod),resid(spmortmod))
```
Differences in mortality rates between species are highly significant and appear to differ in a manner that I expected, based on observing the raw data.
Variance is pretty constant: homoscedastic assumption not violated. Plot level random effect `pl` is doing a good job of capturing some variation in the data. Although, `foresttype` random effect is capturing even more. Correlation of fixed effects is pretty low.

**Estimate species mortality rates:**
```{r}
fixef(spmortmod)
mortpars<- vector(length=length(fixef(spmortmod)))
mortpars[1]<- fixef(spmortmod)[1]
# Extract par estimates from treatment contrasts
mortpars[2:length(mortpars)]<- fixef(spmortmod)[2:length(mortpars)]+mortpars[1] 
# Back-transform logit values on to the original scale
mortrates<- matrix(c(levels(MortDat$species)[c(1,3:17)], 1/(1+(1/exp(mortpars)))), ncol=2) 
mortrates
```
Object `mortrates` stores the estimated mortality rates (proportion of survival) for all species.  
Percentage of survival ranges from 11% to 49% among species.
***
### Possible next steps

Subset data by block. Differences between blocks?

#### Model checking
Obtain R^2 from GLMM using [Nakagawa & Schielzeth's (2012)](http://onlinelibrary.wiley.com/doi/10.1111/j.2041-210x.2012.00261.x/abstract?deniedAccessCustomisedMessage=&userIsAuthenticated=false) method.

Use cross-validation to see how well the parameter estimates generalise.

Fit model to set of data randomly drawn from binomial distribution, instead of survival data, to make sure the estimates are actually representing real world significance, given the data.
